### 2025-11-25 Paper/paper_2025.tex
Updates from Overleaf

    Consolidation in health care remains a forefront concern for regulators in the US, as both horizontal and vertical integration have substantially increased over the last several decades (\cite{levinson2024ten}). Researchers have documented higher prices (\cite{brot2024pays}), and mixed effects on quality (\cite{kessler2000hospital}; \cite{cooper2011does}) as a result of health care becoming more concentrated through mergers and acquisitions. There is significant antitrust scrutiny in this sector, and proposed mergers are regularly denied due to expected harmful effects to competition (\cite{Meyer2022}). Firms in other industries have been shown to engage in methods of informal affiliation besides formal mergers and acquisitions. For example, instances of board overlap in public firms have been cited to violate the Clayton Act due to their anti-competitive nature. Yet, despite concerns of low competition, this type of connectedness among firms has not been given much consideration in health care. In this paper, I document the extent of overlapping board members in nonprofit hospitals in the US, and explore avenues of future research to investigate 
    potential anti-competitive effects of this type of affiliation.

    The presence of overlapping board members across firms, also referred to as board interlock, has been heavily studied in corporate finance, particularly for large publicly traded firms. Board overlap can facilitate indirect collusion on pricing and output strategies, and thus by the Clayton Act of 1914, board overlap among sufficiently large competing firms is illegal. Yet, it is known to exist even among the largest publicly traded firms. Researchers have found that board overlap is correlated with governance practices, performance, investment in research and development, and product differentiation (\cite{cai2014board}; \cite{lamb2016ties}; \cite{geng2021does}). However, it is still unclear whether board overlap really is anti-competitive, as it could also facilitate valuable knowledge sharing and allow multiple firms to gain from expertise of high quality board members. While the Department of Justice has increased scrutiny of board overlap, their focus has remained on tech and private firms (\cite{Morse2023}). Board overlap does occur in health care with instances of organizations denied merger permission due to boards not acting independently (\cite{huberfeld2006tackling}), of hospitals declaring informal affiliations through leadership (\cite{barnett_babcock_2012}), and of life sciences companies with substantial documented overlap (\cite{manjunath2024illegal}). However, little is known about how prevalent board overlap is in the hospital industry, and whether it is correlated with other hospital characteristics.

    In this paper, I form networks of hospital board of director members using Tax Form 990s, which are yearly declarations of the financial state of nonprofits in the US that also include information about who is leading the firm. Extracting these identities and matching to other sources of hospital information, I show that approximately 8\% of the nonprofit hospitals in the sample are affiliated with another hospital in the same market through shared board members, even excluding system or network affiliations. Among hospitals that share board members, two thirds of the hospital pairs are between two general adult hospitals. That is, a majority of the shared board members serve on directly competing hospitals in the same geographic market. In addition, half of the hospital pairs are among one large and one small hospital, and half of the hospital pairs are between two hospitals that do not belong to a system.

    Informed by established research on board overlap and hospital consolidation, I describe categories of hospital behaviors that may be affected by board overlap: price/quality, investment decisions, and differentiation. I construct various measures of behavior in these categories, and show how they are correlated with board overlap for small/large and system/independent hospitals. While the purpose of this paper is not to establish a causal relationship between board overlap and hospital behavior, there are key observable differences in the average quality and investment decisions of hospitals with and without board overlap that motivate future research on the topic.

     This paper contributes to our understanding of board overlap in a setting not previously considered and our understanding of hospital affiliations apart from formal mergers and acquisitions. The findings have implications for policymakers in assessing potential antitrust violations in certain nonprofit settings, and for future research in identifying the causal effects of board overlap in hospitals.

@@

    \section{Documenting Overlapping Board Members in US Hospitals}\label{sec:document}

    \subsection{Data Construction}

   I compile data on nonprofit hospital board of directors from publicly available Tax Form 990s, which all sufficiently large nonprofits must file with the Internal Revenue Service (IRS) each year. These forms contain a section in which nonprofits declare their board, executives, and highest compensated employees. This paper focuses only on the board of director identities. I limit to firms that fill out Schedule H of the tax form, an indicator that the firm operates a hospital. PDF versions of the 990s are made publicly available by Nonprofit Explorer dating back at least 20 years. However, using optical character recognition text extraction methods on these PDFs is both time consuming and can lead to small errors in names extracted. Since I rely heavily on the accuracy of matching individuals, I analyze the data from 2017 onward, when the IRS began publishing XML versions of these files, which ensures accurate names that can be matched across hospitals and time. This also provides a more current picture of the prevalence of board overlap in this industry. I extract all names and positions of individuals listed in the section titled ``Officers, Directors, Trustees, Key Employees, and Five Highest Compensated Employees". I also record information on the hospital that filed the tax form: the Employee Identification Number (EIN), organization name and location, and other characteristics. There are 2,096 hospitals in the initial extraction of firms and identities.

   Next, I match hospitals in the tax form data to hospitals in other publicly available data sources. The only common information about hospitals across these sources is the name and location. Therefore, I use fuzzy string matching methods to create a crosswalk of EINs to identification numbers in the American Hospital Association (AHA) survey data. Specifically, I compute the Jaro-Winkler distance between hospital names in each data set. Then, I record matches with sufficiently similar names and addresses. I have a sample of 1,623 hospitals in the tax data that can be linked to the AHA survey.

    I limit the individuals in the sample to those on the board of directors, excluding executives that also serve on the board. I clean the board member names by removing common titles (Dr., Reverand, etc.), and combining names within the same firm that have slight spelling differences. For example, if John Matthews is a board member of hospital A in 2017, and John Matthew is a board member of hospital A in 2018, I assume these names represent the same person. Finally, to avoid issues in whether first or last names are listed first, I sort each individual's names alphabetically.

    I define a hospital as having an overlapping board member if any of the individuals on their board are also on the board of a hospital in the same hospital referral region (HRR). HRRs are market designations created by Dartmouth Atlas using patient referral data which are meant to capture geographic regions of the average patient's choice set of hospitals. I make this restriction first because board overlap is predicted to matter among competing firms, and hospitals compete for patients within the same geographic region. Additionally, hospital mergers have been shown to have a more significant effect when the entities involved are located geographically close (\cite{cooper2019price}). Therefore, the relevant measure for board overlap in hospitals should be those within the same market.

    Another restriction that I place for two hospitals to share overlapping board members is that they must not be affiliated by system or network. It is common for within-system hospitals to either share common board members or be governed by the same over-arching board. However, systems are already formally affiliated, and this affiliation can dictate many of the behaviors I discuss in Section \ref{sec:hospbehaviors}. Therefore, I rely on variation in board overlap that does not include system affiliation. Thus, any hospitals with board overlap in my sample are otherwise unaffiliated. There still may exist board overlap across systems. That is, I still consider hospital A belonging to System 1 as eligible to share board members with hospital B in System 2. The final sample consists of approximately 1500 nonprofit hospitals from 2017-2022.

@@

    \import{Objects}{all_summarystats.tex}

>>>>

    % Consolidation in health care remains a forefront concern for regulators in the US, as both horizontal and vertical integration have substantially increased over the last several decades (\cite{levinson2024ten}). Researchers have documented higher prices (\cite{brot2024pays}), and mixed effects on quality (\cite{kessler2000hospital}; \cite{cooper2011does}) as a result of health care becoming more concentrated through mergers and acquisitions. There is significant antitrust scrutiny in this sector, and proposed mergers are regularly denied due to expected harmful effects to competition (\cite{Meyer2022}). Firms in other industries have been shown to engage in methods of informal affiliation besides formal mergers and acquisitions. For example, instances of board overlap in public firms have been cited to violate the Clayton Act due to their anti-competitive nature. Yet, despite concerns of low competition, this type of connectedness among firms has not been given much consideration in health care. In this paper, I document the extent of overlapping board members in nonprofit hospitals in the US, and explore avenues of future research to investigate 
    % potential anti-competitive effects of this type of affiliation.

    % The presence of overlapping board members across firms, also referred to as board interlock, has been heavily studied in corporate finance, particularly for large publicly traded firms. Board overlap can facilitate indirect collusion on pricing and output strategies, and thus by the Clayton Act of 1914, board overlap among sufficiently large competing firms is illegal. Yet, it is known to exist even among the largest publicly traded firms. Researchers have found that board overlap is correlated with governance practices, performance, investment in research and development, and product differentiation (\cite{cai2014board}; \cite{lamb2016ties}; \cite{geng2021does}). However, it is still unclear whether board overlap really is anti-competitive, as it could also facilitate valuable knowledge sharing and allow multiple firms to gain from expertise of high quality board members. While the Department of Justice has increased scrutiny of board overlap, their focus has remained on tech and private firms (\cite{Morse2023}). Board overlap does occur in health care with instances of organizations denied merger permission due to boards not acting independently (\cite{huberfeld2006tackling}), of hospitals declaring informal affiliations through leadership (\cite{barnett_babcock_2012}), and of life sciences companies with substantial documented overlap (\cite{manjunath2024illegal}). However, little is known about how prevalent board overlap is in the hospital industry, and whether it is correlated with other hospital characteristics.

    % In this paper, I form networks of hospital board of director members using Tax Form 990s, which are yearly declarations of the financial state of nonprofits in the US that also include information about who is leading the firm. Extracting these identities and matching to other sources of hospital information, I show that approximately 8\% of the nonprofit hospitals in the sample are affiliated with another hospital in the same market through shared board members, even excluding system or network affiliations. Among hospitals that share board members, two thirds of the hospital pairs are between two general adult hospitals. That is, a majority of the shared board members serve on directly competing hospitals in the same geographic market. In addition, half of the hospital pairs are among one large and one small hospital, and half of the hospital pairs are between two hospitals that do not belong to a system.

    % Informed by established research on board overlap and hospital consolidation, I describe categories of hospital behaviors that may be affected by board overlap: price/quality, investment decisions, and differentiation. I construct various measures of behavior in these categories, and show how they are correlated with board overlap for small/large and system/independent hospitals. While the purpose of this paper is not to establish a causal relationship between board overlap and hospital behavior, there are key observable differences in the average quality and investment decisions of hospitals with and without board overlap that motivate future research on the topic.

    %  This paper contributes to our understanding of board overlap in a setting not previously considered and our understanding of hospital affiliations apart from formal mergers and acquisitions. The findings have implications for policymakers in assessing potential antitrust violations in certain nonprofit settings, and for future research in identifying the causal effects of board overlap in hospitals.

@@

    \section{Background/Setting}

    \section{Data Construction}

   I compile data on nonprofit hospital board of directors in years 2017-2022 from publicly available Tax Form 990s, which all sufficiently large nonprofits must file with the Internal Revenue Service (IRS) each year. These forms contain a section in which nonprofits declare their board, executives, and highest compensated employees. This paper focuses only on the board of director identities. I limit to firms that fill out Schedule H of the tax form, an indicator that the firm operates a hospital. I extract all names and positions of individuals listed in the section titled ``Officers, Directors, Trustees, Key Employees, and Five Highest Compensated Employees". I also record information on the hospital that filed the tax form: the Employee Identification Number (EIN), organization name and location, and other characteristics. The initial extraction yields 2,096 nonprofit hospitals.

   Next, I match hospitals in the tax form data to hospitals in other publicly available data sources. The only common information about hospitals across these sources is the name and location. Therefore, I use fuzzy string matching methods to create a crosswalk of EINs to identification numbers in the American Hospital Association (AHA) survey data. Specifically, I compute the Jaro-Winkler distance between hospital names in each data set. Then, I record matches with sufficiently similar names and addresses, yielding a sample of 1,836 matched hospitals.

    I limit the individuals in the sample to those on the board of directors, excluding executives that also serve on the board. I then record self-reported information about the board members in the data, including whether they include a title for medical doctor, nurse, or having a degree in health administration. I also record whether the firm records that they are a non-voting member of the board (though this is very rare). Several steps are necessary to ensure the names are clean and consistent across firms to identify people across hospitals. First, I remove any titles or credentials, and I convert common nicknames to their longer versions for consistency. Second, I combine names within the same firm that have either slight misspelling differences or, if middle names are only sometimes listed, they have at least 2 names in common. For example, if John Lee Matthew is a board member of hospital A in 2017, and John Matthew is a board member of hospital A in 2018, I assume these names represent the same person. I use the gender package in R to predict each individual's gender using historical data (\cite{gender}).

    I create a list of overlapping board member candidates by identifying any names that occur on multiple hospital boards in the same year. However, it is very common for hospitals within a system to largely share the same board. This is not the type of variation I seek to investigate in this paper, since hospitals within systems are already heavily affiliated. Therefore, I define the relevant type of overlap as someone being on the board of multiple hospitals that are otherwise \textit{unaffiliated}. Hospitals within systems are still eligible to have overlap with other hospitals that do not belong to the same system as they do. Further, In the hospital merger literature, mergers have been found to be more significant when the two hospitals involved are located geographically close \cite{cooper2019price}. Therefore, I define two types of overlap: overlap with an otherwise non-affiliated hospital in the same market, or overlap with an otherwise unaffiliated hospital in a different market. All other hospitals have no board overlap with otherwise unaffiliated hospitals. I use the market definitions developed by Dartmouth Atlas, Hospital Referral Regions (HRRs), which are meant to capture geographic regions of the average patient's choice set of hospitals.

    The final sample limitation I make for simplicity is to remove hospitals that gain and then lose overlapping board members in the sample. Since treatment will be turning on overlapping board member, I remove those who "lose" their treatment.

    \subsection{Documenting Overlapping Board Members}

    The full sample of hospitals

@@

    %\import{Objects}{all_summarystats.tex}
